Supplementary Information

### 2021-04-21 doc/manuscript/ForestGEO_treerings.Rmd
"time window" --> "seasonal window"

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically not the first focus for sampling. 
Many worked around the non-linearity issues by sampling at sites with characteristics that result in trees being more climate-limited, though not all [@cook_treering_1982].
Importantly, traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
*Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different time windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

@@

Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the time window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address the following questions and associated hypotheses (Table `r hypotheses`) for a globally distributed tree-ring dataset:
(1) How does interannual climate variation shape annual tree growth (ring width, RW)?
(2) How does growth rate vary with DBH?
(3) How have growth rates changed through time?

@@

Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and time windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate windows across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-time window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site. In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the time window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - time window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for time windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each time window, and (**d**) gives the $\Delta$AIC for each. The time window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

### Most influencial climate drivers {-#Most}
```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW:      (10)
  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
  # non-overlapping: HFt, (1)
#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)
```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential time windows. 
However, the time window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
For eight of 20 site-variable group (i.e., water and temperature) combinations, both both the most influential climate variable and time window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping time windows  (e.g., Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping time windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different time windows and, for precipitation, different variables (PPT and PDF; Figs. `r climwin_HF_W`- `r climwin_HF_T`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
#LDW: precip c June (1 mo), PET current June (1mo) W=T
#HF: PDF pMay (1mo), PET cMar (1mo) W=T
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
```
Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
Optimal time windows most commonly coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.

The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
Optimal time windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same time window (exception: LDW).

@@

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal time windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

In this study, we present a novel robust approach that allows integrative analysis of factors that jointly influence tree growth (Table `r hypotheses`). 
Specifically, it allows us to demonstrate that 
(1) growth is generally simultaneously influenced by temperature and precipitation over different time windows, 
(2) nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
(3) growth -- expressed in any metric -- almost universally varies with DBH (Fig. `r size`), 
and (4) current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.

>>>>

*(reword to ecology focus):*
Dendrochronological studies most commonly focus on linear climate responses to individual climate drivers, thereby missing nonlinearities and additive or interactive climate effects known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020]. 
Nonlinearities in climate sensitivities of tree metabolism and growth have been observed across a wide range of spatio-temporal scales.
Over time frames of seconds to days, photosynthesis and respiration display unimodal relationships to temporal variation in temperature, typically peaking at temperatures reflective of the environment to which the plant is adapted and acclimated [@kumarathunge_acclimation_2019]. 
Across broad climatic gradients, annual forest productivity generally increases with temperature, precipitation, and potential evapotranspiration (PET)  up to a point, after which it plateaus or decreases [@sullivan_longterm_2020; @banburymorgan_global_2021]. 
Filling a critical gap between short-term physiological responses and the global gradients representing millennia of community assembly and species adaptation, the annual growth records of tree rings capture tree growth responses to interannual climatic variation.
Yet, because non-linearities are problematic for reconstructing climate variables [@esper_divergence_2009], systems exhibiting these are typically not the first focus for sampling. 
Many worked around the non-linearity issues by sampling at sites with characteristics that result in trees being more climate-limited, though not all [@cook_treering_1982].
Importantly, traditional analysis methods are designed around first-order linear growth-climate relationships [@fritts_tree_1976]. 
*Dendrochronological studies allowing for nonlinear or threshold responses of tree growth to climate are less common [@cook_climate_1989; @woodhouse_artificial_1999; @tolwinski-ward_bayesian_2013; @tumajer_increasing_2017; @cavin_highest_2017; @ljungqvist_assessing_2020; @rollinson_climate_2021] **one could say these were a lot of citation for it to be uncommon. Perhaps make it more clear what aspects are less common?**, and we therefore know little about the nonlinearities in tree growth responses to interannual variation in climate that occur for trees within forest settings.*
Furthermore, temperature and moisture are known to jointly shape tree growth [@beedlow_importance_2013; @foster_predicting_2016] and forest productivity [e.g., @alexander_relative_2018; @banburymorgan_global_2021], yet growth sensitivity to their additive or interactive effects, potentially operating over different seasonal windows, is not commonly considered [but see @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015], **largely because dendroclimate studies sample in such a way as to isolate one or other of moisture or temperature limitation.**

@@

Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address the following questions and associated hypotheses (Table `r hypotheses`) for a globally distributed tree-ring dataset:
(1) How does interannual climate variation shape annual tree growth (ring width, RW)?
(2) How does growth rate vary with DBH?
(3) How have growth rates changed through time?

@@

Data analysis consisted of two main steps: (1) identifying the primary climate drivers (i.e., variables and seasonal windows over which they are most influential on tree growth), and (2) combining these climate drivers, DBH, and year into a multivariate model (Fig. `r schematic`). 
The analysis was run separately for each site (step 1), site-species combination (step 2), and each response variable estimating different measures of tree growth (RW, BAI, or $\Delta$AGB). 
We note that the decision to identify primary climate drivers at the level of site, as opposed to species, was motivated by the expectation that differences in optimal climate drivers across species in one site would be minimal compared to cross-site differences (Fig. `r SCBI_example`); however, analyses focused on interspecific differences could optimize species-specific climate sensitivity estimates by fitting *climwin* individually by species.

![**Figure 1 | Schematic illustration of the analysis process.** In step 1, the R package *climwin* [@vandepol_identifying_2016] is used to identify the primary climate drivers in water and temperature variable groups for each site, defined as the variable-seasonal window combination that are most strongly correlated to the residual variation around splines fit to trends in growth (here, ring width, RW) for all cores sampled at the site. In step 2, a GLS model is used to produce a combined model with the previously identified drivers, reconstructed DBH, and year.](tables_figures/schematic_figure.png)

@@

We used the *climwin* package in R [@vandepol_identifying_2016] to identify the most important climate variable and the seasonal window over which its effect was strongest for each of two categories of variables: a temperature group ($T_{min}$, $T_{max}$, $T_{mean}$, and PET) and a precipitation group (PPT, PDF).
To remove low-frequency variation that most likely represents responses to non-climatic drivers (*e.g.*, growth and aging of the tree, change in competitive dynamics, atmospheric pollution), we detrended the response variables by fitting penalized thin-plate regression splines in generalized additive models [GAM, functions *gam* and *s* in the R package *mgcv*, @wood_fast_2011] to individual growth records (RW, BAI, or $\Delta$AGB) from each core, and extracted the residual variation for each observation. 
The smoothing parameters were automatically selected by the *gam* function with generalized cross-validation (GCV).
We then used *climwin* to identify the climate variables that most strongly correlated with the individual tree-level residuals of the growth variables, RW, BAI, or $\Delta$AGB, specifying quadratic fits to allow for potential nonlinearities in the climate response.
Within *climwin*, we specified a mixed-effects model in which the fixed effects were the climate variables and the random intercepts were species (when n $\ge$ 3) and core identity (noting that these effects should be minimal given that residuals are centered around zero).
For each climate variable, we ran permutations for all possible combinations of consecutive months within a 15-month period ending near the time of cessation of formation of each annual ring (Table `r site_details`). 
*Climwin* runs all potential models to select the best fit (lowest AIC), and does k-fold cross-validation in its computation of AIC to guard against over-fitting [@vandepol_identifying_2016]. 
For each group of candidate climate variables (water and temperature; Fig. `r schematic`), we selected the variable - seasonal window combination identified by *climwin* as having the lowest AIC as a candidate climate variable for the multivariate models.

@@

![**Figure 2 | Example comparison of climate sensitivity derived via traditional methods (**a**) and our approach (**b-f**).** Example is for the sensitivity of 14 species at SCBI (codes given in Table `r species_table`) to potential evapotranspiration (PET). Panel (**a**) shows a matrix of Pearson correlations between ring-width index and monthly climate variables [produced using the bootRes package in R, @zang_dendroclimatic_2013]. Black rectangle represents the period selected by *climwin* as the most influential window. Panels (**b-d**) give statistics for seasonal windows tested in *climwin*, where window open and close indicate months prior to current August, and cells across the lower diagonal indicate single-month tests (akin to panel **a**). Panels (**b**) and (**c**) give values of linear and quadratic terms for each seasonal window, and (**d**) gives the $\Delta$AIC for each. The seasonal window with the minimum $\Delta$AIC (1-3 months prior to August, or May-July; black circles), was identified as the optimal window. Panel (**e**) shows the correlation of individual-level residuals to PET, with the function fit in *climwin*. Finally, panel (**f**) shows the GLS model output, where PET was a candidate driver variable (along with PPT; DBH not included in this model). Plotted are responses of species for which PET was included in the top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are significant, dashed lines when only one term is significant, and dotted lines when neither is significant. Transparent ribbons indicate 95% confidence intervals. Species names corresponding to the codes are given in Table `r species_table`.](tables_figures/quilt_comparison.png)

@@

### Most influencial climate drivers {-#Most}
```{r , eval=FALSE, echo=FALSE}
## Tallies for the following paragraph (Feb 11)
#sameV_sameW: LDWp, NMt,NMp, NIOt, NIOp, SCBIp,  ZOFp,ZOFt,  (8)
#sameV_diffW:      (10)
  # overlapping: BCNMp, BCNMt, HKKp, SCBIt, LDWt, CBp,CBt , SCp, SCt, (9)
  # non-overlapping: HFt, (1)
#diffV_diffW: HFp (non-overlapping), HKKt (overlapping)
```
At each site, the three metrics of growth (RW, BAI, and $\Delta$AGB) exhibited similar patterns in the direction of response and relative strength of correlation to climate variables across the range of potential seasonal windows. 
However, the seasonal window exhibiting the strongest climatic effect on growth and even the most influential climate variable sometimes differed among the growth metrics (Figs. `r all_climwin`).
For eight of 20 site-variable group (i.e., water and temperature) combinations, both both the most influential climate variable and seasonal window were identical across growth metrics (e.g., Fig. `r climwin_SCBI_W`). 
For nine site-variable group combinations, *climwin* identified the same climate variable and overlapping seasonal windows  (e.g., Fig. `r climwin_SCBI_T`), 
and in one case (at HKK) different variables ($T_{max}$ and $T_{mean}$) were selected with overlapping seasonal windows (Fig. `r climwin_HKK_T`). 
For just two site-variable group combinations (both variable groups at HF, where climate had only weak effects and mixed responses among species in the final models), *climwin* identified completely different seasonal windows and, for precipitation, different variables (PPT and PDF; Figs. `r climwin_HF_W`- `r climwin_HF_T`).
Henceforth, unless otherwise noted, we focus on the climate sensitivities identified using RW as the growth metric and for the full set of cores (i.e., including those for which DBH could not be reconstructed).

@@

###Windows selected: 
# BCNM: precip dry season (12mo), T dry season (2 mo)   W>T
#HKK: PDF dry season + edges (9mo), T wet season (8 mo)  W>T
#SCBI: precip p June - cAug (15 mo), PET current gs (3 mo) W>T
#LDW: precip c June (1 mo), PET current June (1mo) W=T
#HF: PDF pMay (1mo), PET cMar (1mo) W=T
#ZOF: pre pJune-pJul (2mo), TMX Jan-Mar (3mo) W<T
#NIO: ppt pOct-pDec (3mo), TMP pJune (1mo) W>T
#LT: P pNov-cJul (9mo), TMX cMay-Jul (3mo) W>T
#CB: P pJune-cJul(14mo), TMXpAug-pAug(1mo) W>T
#SC: P pJun-pAug(3mo), TMXcJul (1mo) W>T
```
Precipitation amount (PPT) was selected over precipitation frequency (PDF) as the top variable in five of the eight sites with both variables available (but had no significant main effect at one site, NIO), and was the only option at two sites (LT and CB).
Optimal seasonal windows most commonly coincided at least partially with months of active growth in the current year (Fig. `r clim_sens`; Table `r site_details`):  year-round in the tropics (BCNM and HKK) or late spring/ summer in the extratropics (n= 5 of 8).
In the tropics, the long time-windows over which precipitation was influential (12 mo at BCNM, 9 mo at HKK) also included the majority (BCNM) or all (HKK) of the dry season.
In the extratropics, optimal windows at three sites included the current growing season and extended back to the previous fall (LT, CB) or summer (SCBI), whereas they were limited to the current spring and early summer at LDW.
At three sites (HF, ZOF, and SC), precipitation of the previous growing season was the most influential variable.

The most commonly selected variables within the temperature group were $T_{max}$ and PET, which were identified by *climwin* as the top temperature-related driver at six and three of the 10 sites, respectively, noting that PET was not available for two sites. 
$T_{min}$ was identified as the top driver at BCNM, where it's effects were only marginally significant for one species (Fig. `r clim_sens`), and $T_{mean}$ was never selected as the top driver.
Optimal seasonal windows for temperature tended to be shorter than those of precipitation (Table `r hypotheses`).
They most commonly occurred during the current growing season (n= 5 of 10 sites), but there were cases where optimal windows occurred during the preceding dry season (BCNM), late winter/early spring (HF, ZOF), or the previous growing season (NIO, CB).
Temperature and precipitation variables were rarely influential over the same seasonal window (exception: LDW).

@@

A temperature variable was included in the best model for all but eight site-species combinations, with predominantly negative responses, particularly at the higher end of the temperature range (`r round(percent_temperature_responses_1st_order_negative + percent_temperature_responses_1st_order_pos_2nd_order_neg)`%; `r round(percent_temperature_responses_1st_order_negative)`% with negative first-order term, `r round(percent_temperature_responses_1st_order_pos_2nd_order_neg)`% with positive first-order term but negative second-order term; Fig. `r clim_sens`).
Within the tropics, there was minimal effect of temperature at BCNM and a negative effect of wet season $T_{max}$ for three of four species at HKK.
For temperate sites with optimal seasonal windows covering the current and/or past growing season, responses were universally negative (i.e., negative first-order linear or unimodal, peaking at temperatures lower than the long-term mean). 
In contrast, there were positive effects of Jan-March $T_{max}$ for all three species at ZOF and of March PET for *Tsuga canadensis* at HF, the latter contrasting with a negative response of the three deciduous species analyzed at HF (Fig. `r clim_sens`).
At the highest-latitude site (SC), which has undergone rapid warming and permafrost melt, *Picea mariana* responded positively (but with wide 95% CI on the slope) to temperature over the full analysis period (1903-2013); however, responses were predominately positive prior to 1970 and predominantly negative afterwards (Fig. `r SC_1970break`).

@@

In this study, we present a novel robust approach that allows integrative analysis of factors that jointly influence tree growth (Table `r hypotheses`). 
Specifically, it allows us to demonstrate that 
(1) growth is generally simultaneously influenced by temperature and precipitation over different seasonal windows, 
(2) nonlinear climate responses and interactive effects of climate with DBH are common (Figs. `r clim_sens` and `r interactions_fig`, respectively),
(3) growth -- expressed in any metric -- almost universally varies with DBH (Fig. `r size`), 
and (4) current or recently living trees commonly show growth declines in response to some combination of stand dynamics and environmental change (Fig. `r year_fig`).
While traditional dendrochronology methods, particularly those focused on climate reconstruction, generally sample and analyze data to minimize many of these effects, they are critical for understanding forest productivity in an era of global change.
As global change pressures intensify and the need to understand changing forest dynamics becomes increasingly urgent [@thom_impact_2017; @mcdowell_pervasive_2020], we expect that this approach will prove valuable to understanding drivers of tree growth and forest change.
